Quantizing the Bosonic String on a Loop Quantum Gravity Background

### 2024-08-28 temp/lqg-strings.tex
Changed title to reflect conclusion, revised abstract

Following this, work by Rovelli and Smolin \cite{Rovelli1993Area,Rovelli1994Discreteness} demonstrated a remarkable feature of this theory, which came to be known as ``Loop Quantum Gravity'', was that one could construct quantum operators for geometric quantities such as areas of two-dimensional surfaces and volumes of three-dimensional regions. Moreover, these operators could be diagonalized exactly in the spin-network basis and a discrete spectrum for quantum of area and quantum of volume could be derived. This was the first time that physicists had discovered the ``atoms of space'' or the ``quanta of geometry'' in the true sense of the expression, in a theory of quantum gravity. One of the early triumphs of this formalism, which has unfortunately not received enough attention over the years by the larger high energy theory community, was the calculation of the entropy of a Schwarzschild black hole \cite{Rovelli1996Black} in a fully quantum mechanical framework and showing that it matched the expected expression $S \sim A/4$ for the Bekenstein-Hawking entropy.

One key feature of the discreteness of spacetime itself is that the Ashtekar connection $A^{i} _{a}$ \footnote{Here, the index convention followed is that lowercase $i , j , k$ refer to 3-dimensional flat internal ($\mf{su}(2)$) indices and lowercase $a , b , c$ refer to 3-dimensional curved spacetime indices} can no longer be a continuous function of spacetime, rather it becomes a \emph{distribution}, assuming a constant value on one patch of space, then another value at another patch, etc \cite{Bodendorfer-LQG,Rovelli2015Covariant}. The consequence of this non-continuous distributional nature of the connection is that the triad (or Ashtekar electric field) operator, which formally would be
\begin{equation} \label{eqn:triad-operator}
	\Hat{\widetilde{E}}{}^{a} _{i} = - 8 \pi i \hbar G \frac{\delta}{\delta A^{i} _{a}}
\end{equation}
becomes ill-defined since $A^{i} _{a}$ is non-continuous (in particular, at the edges of a polyhedra it changes abruptly), so the triad needs to be \emph{smeared} over a surface where $A^{i} _{a}$ is constant
\begin{equation} \label{eqn:triad-flux-operator}
	\Hat{E}_{i} (S) = - 8 \pi i \hbar G \int _{S} \diff S_{a}  \frac{\delta}{\delta A^{i} _{a}} ,
\end{equation}
where $\diff S_{a} = \diff \sigma ^{b} \wedge \diff \sigma ^{c} \varepsilon _{a b c}$. This naturally leads to the \emph{area operator} of said surface
\begin{equation} \label{eqn:area-operator}
	\Hat{A} (S) = - 8 \pi i l_{pl} ^2 \beta \int _{S} \sqrt{\diff S_{a} \diff S_{b} \frac{\delta}{\delta A^{i} _{a}} \frac{\delta}{\delta A_{b i}}} .
\end{equation}
Notably, since $A^{i} _{a}$ is a $\mf{su}(2)$-valued connection, in the basis which diagonalises the area operator (the \emph{spin network} basis \cite{Bodendorfer-LQG}) $\ket{\Tr (h^{(j)} _{c} (A))}$, the eigenvalues of the area operator are just proportional to \emph{angular momentum} eigenvalues
\begin{equation} \label{eqn:area-eigenvalues}
	\Hat{A} \ket{\Tr (h^{(j)} _{c} (A))} = 8 \pi l_{pl} ^2 \beta \sqrt{j (j + 1)} \ket{\Tr (h^{(j)} _{c} (A))} ,
\end{equation}
with $j$ being the representation label of $\mf{su}(2)$ along the spin network edge piercing the surface under consideration. Thus, spin network edges carry angular momenta.

@@

In \autoref{sec:emergent-dimensions} we complement Afshordi's work \cite{Afshordi2014Emergent} with insights from LQG, showing that LQG quantum geometry naturally constrains most of the free parameters of said model, leading to a more concrete theory. We also suggest a possible consequence for the parameters that remain unconstrained and a generalization of the model.

@@

The equations of motion for the world-sheet metric are completely unchanged by this correction,
\begin{equation} \label{eqn:vary-wrt-g}
	\frac{\delta S_{I A P}}{\delta g^{c d}} = 0
\end{equation}
which leads to the stress-energy tensor for the worldsheet:
\begin{multline} \label{eqn:inverse-area-em-tensor}
	T_{c d} := \partial _{c} X^{\mu} \partial _{d} X^{\nu} G_{\mu \nu} - \\ - \frac{1}{2} g_{c d} g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} G_{\mu \nu} = 0 ,
\end{multline}
from which we can extract the metric $g_{c d}$ as
\begin{equation} \label{eqn:ws-metric}
	g_{c d} = 2 f \partial _{c} X^{\mu} \partial _{d} X^{\nu} G_{\mu \nu} ,
\end{equation}
where the function $f$ is given by
\begin{equation} \label{eqn:inverse-area-f}
	\frac{1}{f} = g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} G_{\mu \nu}
\end{equation}
just like the regular ST literature, thus the constraints are unchanged by our correction.

As for variation \wrt the embbeding fields $X^{\lambda}$, we have
\begin{equation} \label{eqn:vary-wrt-X}
	\frac{\delta S_{I A P}}{\delta X^{\lambda}} = 0
\end{equation}
\begin{multline} \label{eqn:X-eom-inverse-area}
	F^{-} \sqrt{- g} g^{a b} \partial _{a} \partial _{b} X^{\nu} G_{\lambda \nu} + \partial _{a} \left( F^{-} \sqrt{- g} g^{a b} \right) \partial _{b} X^{\nu} G_{\lambda \nu} = \\ = \frac{1}{2} F^{-} \sqrt{- g} g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} \partial _{\lambda} G_{\mu \nu} ,
\end{multline}
where
\begin{equation} \label{eqn:F-minus-def}
	F^{-} \equiv 1 - \frac{k \Delta}{2 (- g)} .
\end{equation}
Since all 2-dimensional metrics are conformally flat \cite{Schottenloher2008A-Mathematical}, we use conformal gauge $g_{a b} = \e ^{\phi (x)} \eta _{a b}$ and assume flat background $G_{\mu \nu} = \eta _{\mu \nu}$, to simplify eq (\ref{eqn:X-eom-inverse-area}) into
\begin{equation} \label{eqn:X-eom-simplified}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \frac{1}{F^{-}} \eta ^{a b} \partial _{a} F^{-} \partial _{b} X^{\mu} = 0 .
\end{equation}
By expanding the definition of $F^{-}$ we get
\begin{equation} \label{eqn:coupling-X-phi}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \eta ^{a b} \frac{\left( 2 \del _{a} \phi - \del _{a} \ln (\Delta) \right)}{\dfrac{2 \e ^{2 \phi}}{k \Delta} - 1} \del _{b} X^{\mu} = 0 ,
\end{equation}
revealing a wave equation sourced by a coupling between the WS conformal factor, quantum of area field $\Delta$ and the embedding fields much akin to the dilaton field \cite{Balthazar-t-backgrounds}. We'll come back to this in \autoref{sec:emergent-dimensions}.

@@

We now seek to quantize the Klein-Gordon string by means of covariant quantization, in a manner much similar to those found in \cite{Tong2009Lectures,Green2012Superstring}. The string coordinates $X^{\mu}$ and it's conjugate momentum $\Pi ^{\nu}$ are turned into operators $\Hat{X}^{\mu}$ and $\Hat{\Pi}^{\nu}$, with the Poisson brackets $\{ . , . \}$ being replaced by commutators $\{ . , . \} \rightarrow - i [ . , . ]$:
\begin{subequations}
	\begin{align}
		\left[ \Hat{X}^{\mu} (\tau , \sigma) , \Hat{\Pi}^{\nu} (\tau , \sigma ') \right] & = i \delta (\sigma - \sigma ') \eta ^{\mu \nu} \label{eqn:X-P-commutator}                                              \\
		\left[ \Hat{X}^{\mu} , \Hat{X}^{\nu} \right]                                     & = 0 = \left[ \Hat{\Pi}^{\mu} , \Hat{\Pi}^{\nu} \right] \label{eqn:X-X-P-P-commutators}                                 \\
		\left[ \Hat{a}^{\mu} _{n} , \Hat{a}^{\nu} _{p} \right]                           & = 0 = \left[ (\Hat{a}^{\mu} _{n})^{\dagger} , (\Hat{a}^{\nu} _{p})^{\dagger} \right] \label{eqn:a-a-a*-a*-commutators} \\
		\left[ \Hat{a}^{\mu} _{n} , (\Hat{a}^{\nu} _{p})^{\dagger} \right]               & = \delta _{|n| , |p|} \eta ^{\mu \nu} \label{eqn:a-a*-commutator}                                                      \\
		\left[ \Hat{x}^{\mu} , \Hat{x}^{\nu} \right]                                     & = 0 = \left[ \Hat{p}^{\mu} , \Hat{p}^{\nu} \right] \label{eqn:x-x-p-p-commutators}                                     \\
		\left[ \Hat{x}^{\mu} , \Hat{p}^{\nu} \right]                                     & = i \eta ^{\mu \nu} \label{eqn:x-p-commutator}
	\end{align}
\end{subequations}

@@

Let us make one crucial observation here. In the expression \eqref{eqn:ng-area-corrected} we originally took $\Delta$ to a constant representing the smallest quantum of area. It is only \emph{later}, by demanding consistency of the area-corrected Nambu-Goto and Polaykov actions that we realise that $\Delta \equiv \Delta_{ab}(x) $ is in fact an antisymmetric worldsheet two-form!

After exploring the equations of motion of the area corrected string, we found that said equations can be recast as a set of Klein-Gordon equations, which for a first analysis we assumed to have constant coefficient $\mu$, however that need not be the case. The solution of these equations and their consequences for the constraints lead us to a modified \textquotedblleft Virasoro-Klein-Gordon" algebra, where the algebra modes have two indices, which in turn resulted in a truncated solution. This also had the effect of not having distinct left and right moving modes, thus there was no level-matching prohibiting spin-1 states for the closed string, which we postulated to maybe be related to the Ashtekar variables of the background geometry. The quantisation of the KG string and the Virasoro-Klein-Gordon algebra presented an anomaly-free algebra, no restrictions on the background dimensionality and no massless spin-1 or spin-2 excitations at the effective level and no negative-norm states in physical processes, with the usual $G_{\mu \nu} , B_{\mu \nu} , \Phi$ fields as part of the spin-2 spectrum.

We also briefly explored an alternative approach to incorporate LQG quantum geometric effects into string theory by considering Afshordi's ``emergent dimensions’’ model \cite{Afshordi2014Emergent} with some key insights from LQG complementing said model, creating a more concrete picture of the mechanism behind the emerging dimensions.

>>>>

Following this, work by Rovelli and Smolin \cite{Rovelli1993Area,Rovelli1994Discreteness} demonstrated a remarkable feature of this theory, which came to be known as ``Loop Quantum Gravity'', was that one could construct quantum operators for geometric quantities such as areas of two dimensional surfaces and volumes of three dimensional regions. Moreover, these operators could be diagonlized exactly in the spin-network basis and a discrete spectrum for quantum of area and quantum of volume could be derived. This was the first time that physicists had discovered the ``atoms of space'' or the ``quanta of geometry'' in the true sense of the expression, in a theory of quantum gravity. One of the early triumphs of this formalism, which has unfortunately not received enough attention over the years by the larger high energy theory community, was the calculation of the entropy of a Schwarzschild black hole \cite{Rovelli1996Black} in a fully quantum mechanical framework and showing that it matched the expected expression $S \sim A/4$ for the Bekenstein-Hawking entropy.

@@

In \autoref{sec:emergent-dimensions} we complement Afshordi's work \cite{Afshordi2014emergent} with insights from LQG, showing that LQG quantum geometry naturally constrains most of the free parameters of said model, leading to a more concrete theory. We also suggest a possible consequence for the parameters that remain unconstrained and a generalization of the model.

@@

The equations of motion for the world-sheet metric are completely unchanged by this correction,
\begin{equation} \label{eqn:vary-wrt-g}
	\frac{\delta S_{I A P}}{\delta g^{c d}} = 0
\end{equation}
which leads to the stress-energy tensor for the worldsheet:
\begin{multline} \label{eqn:inverse-area-em-tensor}
	T_{c d} := \partial _{c} X^{\mu} \partial _{d} X^{\nu} G_{\mu \nu} - \\ - \frac{1}{2} g_{c d} g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} G_{\mu \nu} = 0 ,
\end{multline}
from which we can extract the metric $g_{c d}$ as
\begin{equation} \label{eqn:ws-metric-v2}
	g_{c d} = 2 f \partial _{c} X^{\mu} \partial _{d} X^{\nu} G_{\mu \nu} ,
\end{equation}
where the function $f$ is given by
\begin{equation} \label{eqn:inverse-area-f}
	\frac{1}{f} = g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} G_{\mu \nu}
\end{equation}
just like the regular string theory literature, thus the constraints are unchanged by our correction.

As for variation \wrt the embbeding fields $X^{\lambda}$, we have
\begin{equation} \label{eqn:vary-wrt-X}
	\frac{\delta S_{I A P}}{\delta X^{\lambda}} = 0
\end{equation}
\begin{multline} \label{eqn:X-eom-inverse-area}
	F^{-} \sqrt{- g} g^{a b} \partial _{a} \partial _{b} X^{\nu} G_{\lambda \nu} + \partial _{a} \left( F^{-} \sqrt{- g} g^{a b} \right) \partial _{b} X^{\nu} G_{\lambda \nu} = \\ = \frac{1}{2} F^{-} \sqrt{- g} g^{a b} \partial _{a} X^{\mu} \partial _{b} X^{\nu} \partial _{\lambda} G_{\mu \nu} ,
\end{multline}
where
\begin{equation} \label{eqn:F-minus-def}
	F^{-} \equiv 1 - \frac{k \Delta}{2 (- g)} .
\end{equation}
Since all 2-dimensional metrics are conformally flat \cite{Schottenloher2008A-Mathematical}, we use conformal gauge $g_{a b} = \e ^{\phi (x)} \eta _{a b}$ and assume flat background $G_{\mu \nu} = \eta _{\mu \nu}$, to simplify eq (\ref{eqn:X-eom-inverse-area}) into
\begin{equation} \label{eqn:X-eom-simplified}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \frac{1}{F^{-}} \eta ^{a b} \partial _{a} F^{-} \partial _{b} X^{\mu} = 0 .
\end{equation}
By expanding the definition of $F^{-}$ we get
\begin{equation} \label{eqn:coupling-X-phi}
	\eta ^{a b} \partial _{a} \partial _{b} X^{\mu} + \eta ^{a b} \frac{\left( 2 \del _{a} \phi - \del _{a} \ln (\Delta) \right)}{\dfrac{2 \e ^{2 \phi}}{k \Delta} - 1} \del _{b} X^{\mu} = 0 ,
\end{equation}
revealing a wave equation sourced by a coupling between the WS conformal factor, quantum of area field $\Delta$ and the embedding fields much akin to the dilaton field \cite{Balthazar-t-backgrounds}. We will come back to this fact in \autoref{sec:emergent-dimensions}.

@@

We now seek to quantize the Klein-Gordon string by means of covariant canonical quantization, in a manner much similar to those found in (\cite{Tong2009Lectures},\cite{Green2012Superstring}). The string coordinates $X^{\mu}$ and it's conjugate momentum $\Pi ^{\nu}$ are turned into operators $\Hat{X}^{\mu}$ and $\Hat{\Pi}^{\nu}$, with the Poisson brackets $\{ . , . \}$ being replaced by commutators $\{ . , . \} \rightarrow - i [ . , . ]$:
\begin{subequations}
	\begin{align}
		\left[ \Hat{X}^{\mu} (\tau , \sigma) , \Hat{\Pi}^{\nu} (\tau , \sigma ') \right] & = i \delta (\sigma - \sigma ') \eta ^{\mu \nu} \label{eqn:X-P-commutator}                                              \\
		\left[ \Hat{X}^{\mu} , \Hat{X}^{\nu} \right]                                     & = 0 = \left[ \Hat{\Pi}^{\mu} , \Hat{\Pi}^{\nu} \right] \label{eqn:X-X-P-P-commutators}                                 \\
		\left[ \Hat{a}^{\mu} _{n} , \Hat{a}^{\nu} _{p} \right]                           & = 0 = \left[ (\Hat{a}^{\mu} _{n})^{\dagger} , (\Hat{a}^{\nu} _{p})^{\dagger} \right] \label{eqn:a-a-a*-a*-commutators} \\
		\left[ \Hat{a}^{\mu} _{n} , (\Hat{a}^{\nu} _{p})^{\dagger} \right]               & = \delta _{|n| , |p|} \eta ^{\mu \nu} \label{eqn:a-a*-commutator}                                                      \\
		\left[ \Hat{x}^{\mu} , \Hat{x}^{\nu} \right]                                     & = 0 = \left[ \Hat{p}^{\mu} , \Hat{p}^{\nu} \right] \label{eqn:x-x-p-p-commutators}                                     \\
		\left[ \Hat{x}^{\mu} , \Hat{p}^{\nu} \right]                                     & = i \eta ^{\mu \nu} \label{eqn:x-p-commutator}
	\end{align}
\end{subequations}

@@

Let us make one crucial observation here. In the expression \eqref{eqn:ng-area-corrected} we originally took $\Delta$ to a constant representing the smallest quantum of area. It is only \emph{later}, in \autoref{sec:inverse-metric-Polyakov}, by demanding consistency of the area-corrected Nambu-Goto and Polaykov actions that we realise that $\Delta \equiv \Delta_{ab}(x) $ is in fact an antisymmetric worldsheet two-form \eqref{eqn:Delta-b-relation}!

We then proceeded to explore the equations of motion in \autoref{sec:quantum-kg-string} resulting from the Polyakov form of the area-corrected string action . We found that the constraint algebra must be modified due to the existence of a minimal area and the resulting modification gives us \emph{two-index} Virasoro generators. We found the algebra of generators to be anomaly free implying that the area-corrected string action is consistent in any number of spacetime dimensions.

We then explored the possible relation between our model and the ``emergent dimensions'' model put forward by Afshordi in \cite{Afshordi2014Emergent} in \autoref{sec:emergent-dimensions}.
